Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2022-02-04 _paper.rmd
extended conclusion.

---
title: "Exploring Local Explanations of Non-linear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
  #- name: "Nicholas Spyrison"
  #   address: |
  #     | ORCiD: 0000-0002-8417-0212
  #   email: \email{nicholas.spyrison@monash.edu}
  #   orcid: 0000-0002-8417-0212
  #- name: "Dianne Cook"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 000-0002-3813-7155
  #   orcid: 000-0002-3813-7155
  #- name: "Kimbal Marriott"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 0000-0002-9813-0377
  #   orcid: 0000-0002-9813-0377
abstract: |
 The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction with a point estimate of the linear feature importance in the vicinity of each instance. These can be considered linear projections and can be further explored interactively to understand better the interaction between features used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any instance and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response features. The methods are implemented in the __R__ package __cheem__, available on CRAN.
#documentclass: jmlr
#classoption: article
output:
  ## pdf_document: requires \ref{} 
  #pdf_document:
  ## bookdown::pdf_document2: allows \@ref()
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    number_sections: yes
    fig_caption: yes
    #includes:
    #  in_header: template/jmlr2e.sty
bibliography: cheem_paper.bib
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
  - \usepackage{setspace} ## for title page spacing
  - \usepackage{hyperref} ## for all sorts of linking
---

@@

<!-- Introduce explanatory vs predictive modeling -->
There are different reasons and emphases to fit a model. @breiman_statistical_2001, reiterated by @shmueli_explain_2010, taxonomize modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose, while _predictive_ modeling focuses more on the predictions of out-of-sample instances. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy [@oneil_weapons_2016; @kodiyan_overview_2019]. However, the loss of interpretation presents a challenge.

@@

<!-- Interpretability & data drift -->
Another concern is data drift, a shift in support or domain of the explanatory features (variable or predictors). Non-linear models are typically more sensitive and do not extrapolate well outside the training data domain. Better interpretability of the model means more transparency where models' predictions may be plausible or completely unreliable.

@@

<!-- Reminder of local explanation -->
Consider a highly non-linear model. It can be hard to determine whether small changes in a feature's value will make a class prediction change group or identify which features contribute to an extreme residual. Local explanations shed light on these situations by approximating linear feature importance in the vicinity of a single instance.

@@

To explore the local explanations, an ensemble of plots [@unwin_ensemble_2018] is provided, called the _cheem viewer_. There are two primary plots: the **global view** to give the context of all of the SHAP values, and the **radial tour view** to explore the local explanations with user-controlled rotation. In addition, there are numerous user inputs, including feature selection for the radial tour and instance selection for making comparisons. There are different plots used for categorical and quantitative response. Figures \@ref(fig:classificationcase) and \@ref(fig:regressioncase) contain screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

We have illustrated this method on random forest models using the tree SHAP local explanation, while it could be generally used with any compatible model-explanation pairing. We apply it to the classification and regression tasks. We have created an open-source __R__ package __cheem__, available on [CRAN](https://CRAN.R-project.org/package=cheem), to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided, or upload your data after preprocessing.

>>>>

---
title: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
  #- name: "Nicholas Spyrison"
  #   address: |
  #     | ORCiD: 0000-0002-8417-0212
  #   email: \email{nicholas.spyrison@monash.edu}
  #   orcid: 0000-0002-8417-0212
  #- name: "Dianne Cook"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 000-0002-3813-7155
  #   orcid: 000-0002-3813-7155
  #- name: "Kimbal Marriott"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 0000-0002-9813-0377
  #   orcid: 0000-0002-9813-0377
abstract: |
 The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction with a point estimate of the linear feature importance in the vicinity of each instance. These can be considered linear projections and can be further explored interactively to understand better the interaction between features used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any instance and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response features. The methods are implemented in the __R__ package __cheem__, available on CRAN.
#documentclass: jmlr
#classoption: article
output:
  ## pdf_document: requires \ref{} 
  #pdf_document:
  ## bookdown::pdf_document2: allows \@ref()
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    number_sections: yes
    fig_caption: yes
    #includes:
    #  in_header: template/jmlr2e.sty
bibliography: cheem_paper.bib
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
  - \usepackage{setspace} ## for title page spacing
  - \usepackage{hyperref} ## for all sorts of linking
---

@@

<!-- Introduce explanatory vs predictive modeling -->
There are different reasons and emphases to fit a model. @breiman_statistical_2001, reiterated by @shmueli_explain_2010, taxonomize modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose, while _predictive_ modeling focuses more on the predictions of out-of-sample instances. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate nonlinear models. The use of black-box models is becoming increasingly common, but not without their share of controversy [@oneil_weapons_2016; @kodiyan_overview_2019]. However, the loss of interpretation presents a challenge.

@@

<!-- Interpretability & data drift -->
Another concern is data drift, a shift in support or domain of the explanatory features (variable or predictors). Nonlinear models are typically more sensitive and do not extrapolate well outside the training data domain. Better interpretability of the model means more transparency where models' predictions may be plausible or completely unreliable.

@@

<!-- Reminder of local explanation -->
Consider a highly nonlinear model. It can be hard to determine whether small changes in a feature's value will make a class prediction change group or identify which features contribute to an extreme residual. Local explanations shed light on these situations by approximating linear feature importance in the vicinity of a single instance.

@@

To explore the local explanations, an ensemble of plots [@unwin_ensemble_2018] is provided, called the _cheem viewer_. There are two primary plots: the __global view__ to give the context of all of the SHAP values, and the __radial tour view__ to explore the local explanations with user-controlled rotation. In addition, there are numerous user inputs, including feature selection for the radial tour and instance selection for making comparisons. There are different plots used for categorical and quantitative response. Figures \@ref(fig:classificationcase) and \@ref(fig:regressioncase) contain screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

We have illustrated this method on random forest models using the tree SHAP local explanation. While this analysis could be generally used with any compatible model-explanation pairing. We have created an open-source __R__ package __cheem__, available on [CRAN](https://CRAN.R-project.org/package=cheem), to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided, or upload your data after preprocessing.

One limitation of the __cheem__ packages is that it currently only calculates tree SHAP for tree-based models supported by __treeshap__ (tree based models from __randomForest__, __ranger__, __gbm__, __xgboost__, __lightgbm__, or __catboost__). The SHAP and oscillation explainations could be added with use of the `DALEX::explain()` seems to be an excellent direction to extend [@biecek_dalex_2018; @biecek_explanatory_2021]. Tree SHAP was selected because of its reduced computational complexity; processing runtime of the explanation will be a looming concern.

Attribution space does tend to be more separable than data space at least in the first couple principal components. Different statistics should be explored to convey a clearer understanding what structure the explanation distinguishes. It is curious that the attribution space the same dimensionality of the data with a seemingly better separation. It sounds like a candidate to fit another model. However, such an attempt should want to be vigilant not to over-fit the training data.

### 2022-02-04 _paper.tex
extended conclusion.

  pdftitle={Exploring Local Explanations of Non-linear Models Using Animated Linear Projections},

@@

\title{Exploring Local Explanations of Non-linear Models Using Animated Linear Projections}
\author{}
\date{\vspace{-2.5em}}

@@

There are different reasons and emphases to fit a model. Breiman (2001), reiterated by Shmueli (2010), taxonomize modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose, while \emph{predictive} modeling focuses more on the predictions of out-of-sample instances. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy (O'Neil 2016; Kodiyan 2019). However, the loss of interpretation presents a challenge.

@@

Another concern is data drift, a shift in support or domain of the explanatory features (variable or predictors). Non-linear models are typically more sensitive and do not extrapolate well outside the training data domain. Better interpretability of the model means more transparency where models' predictions may be plausible or completely unreliable.

@@

Consider a highly non-linear model. It can be hard to determine whether small changes in a feature's value will make a class prediction change group or identify which features contribute to an extreme residual. Local explanations shed light on these situations by approximating linear feature importance in the vicinity of a single instance.

@@

We have illustrated this method on random forest models using the tree SHAP local explanation, while it could be generally used with any compatible model-explanation pairing. We apply it to the classification and regression tasks. We have created an open-source \textbf{R} package \textbf{cheem}, available on \href{https://CRAN.R-project.org/package=cheem}{CRAN}, to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided, or upload your data after preprocessing.

>>>>

  pdftitle={Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections},

@@

\title{Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections}
\author{}
\date{\vspace{-2.5em}}

@@

There are different reasons and emphases to fit a model. Breiman (2001), reiterated by Shmueli (2010), taxonomize modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose, while \emph{predictive} modeling focuses more on the predictions of out-of-sample instances. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate nonlinear models. The use of black-box models is becoming increasingly common, but not without their share of controversy (O'Neil 2016; Kodiyan 2019). However, the loss of interpretation presents a challenge.

@@

Another concern is data drift, a shift in support or domain of the explanatory features (variable or predictors). Nonlinear models are typically more sensitive and do not extrapolate well outside the training data domain. Better interpretability of the model means more transparency where models' predictions may be plausible or completely unreliable.

@@

Consider a highly nonlinear model. It can be hard to determine whether small changes in a feature's value will make a class prediction change group or identify which features contribute to an extreme residual. Local explanations shed light on these situations by approximating linear feature importance in the vicinity of a single instance.

@@

We have illustrated this method on random forest models using the tree SHAP local explanation. While this analysis could be generally used with any compatible model-explanation pairing. We have created an open-source \textbf{R} package \textbf{cheem}, available on \href{https://CRAN.R-project.org/package=cheem}{CRAN}, to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided, or upload your data after preprocessing.

One limitation of the \textbf{cheem} packages is that it currently only calculates tree SHAP for tree-based models supported by \textbf{treeshap} (tree based models from \textbf{randomForest}, \textbf{ranger}, \textbf{gbm}, \textbf{xgboost}, \textbf{lightgbm}, or \textbf{catboost}). The SHAP and oscillation explainations could be added with use of the \texttt{DALEX::explain()} seems to be an excellent direction to extend (Biecek 2018; Biecek and Burzykowski 2021). Tree SHAP was selected because of its reduced computational complexity; processing runtime of the explanation will be a looming concern.

Attribution space does tend to be more separable than data space at least in the first couple principal components. Different statistics should be explored to convey a clearer understanding what structure the explanation distinguishes. It is curious that the attribution space the same dimensionality of the data with a seemingly better separation. It sounds like a candidate to fit another model. However, such an attempt should want to be vigilant not to over-fit the training data.
